\section{Running Anything on the BEAM}

Due to the Erlang/OTP infrastructure, there are a number of possible
code generation targets, if we would like to run our code on the BEAM.

To start, there is \emph{Erlang source code} (usual file-extension:
\texttt{.erl}) and \emph{BEAM bytecode} (usual file-extension:
\texttt{.beam}). Neither of these have up-to-date specification
documents\footnote{The last Erlang language specification was written
up in 1999\cite{erlang:1999:spec}. It attained only a draft form, and
is today considered rather outdated.  See also
\url{https://github.com/erlang/spec}.}, let alone any formal
specifications. However, due to Erlang being a fairly popular
programming language\footnote{See Appendix
\longref{app:popularity-of-erlang} for an analysis of the popularity
of Erlang.}, there is a wealth of literature on programming
Erlang\footnote{See also \url{https://www.erlang.org/docs}.}.  In
general however, the best specification of both, is the Erlang/OTP
implementation itself\footnote{See Appendix
\longref{app:overview-of-erlang-source} for an overview of the
Erlang/OTP source code.}.

Erlang source code and BEAM bytecode are a world apart. The first is a
high-level functional programming language. The second is a low-level
bytecode for a register-based virtual machine. To transform one into
the other, the Erlang/OTP compiler performs a number of
transformations and optimisations; passing over the Erlang module
multiple times.

If we generate Erlang source code, we must make sure to efficiently
map our (possibly high-level) model to the high-level Erlang
programming model. If we instead generate BEAM bytecode, we miss out
on the optimisations the Erlang/OTP compiler could have performed for
us.

Furthermore, the Erlang/OTP compiler takes an Erlang module through a
number of intermediate representations. Some of these representations
are serializable---the compiler can both output, and consume files in
such formats. The serializable intermediate representations are
\emph{Core Erlang} (usual file-extension: \texttt{.core}) and
\emph{BEAM assembly} (usual file-extension: \texttt{.S}). As with
Erlang source code, and BEAM bytecode, these do not have up-to-date
specifications, beyond the Erlang/OTP implementation itself.

All the serializable representations along a happy path through the
Erlang/OTP compiler are illustrated in Figure
\ref{fig:serializable-representations}. Generating code in any of
these formats, would allow us to run code on the BEAM, by leveraging
the file-based API of the Erlang/OTP compiler only. 

\NewDocumentCommand\optimising{}{$\circlearrowright$}

\begin{figure}[ht!]
\centering
\tikzstyle{block} = [rectangle, draw, 
    text width=4.5em, text centered, rounded corners,
    rectangle split, rectangle split parts=2,
    rectangle split draw splits=true,
    node distance=7em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{wrapper} = [rectangle, draw, dashed, rounded corners]
\tikzstyle{optimising} = [line width=0.2em]

\begin{tikzpicture}[auto]
\node (source) [block] {
  \nodepart{one} Erlang source code
  \nodepart{two} \texttt{.erl}
};
\node (core) [block, right of=source] {
  \nodepart{one} Core Erlang\\ \optimising{}
  \nodepart{two} \texttt{.core}
};
\node (asm) [block, right of=core] {
  \nodepart{one} BEAM assembly\\ \optimising{}
  \nodepart{two} \texttt{.S}
};
\node (bin) [block, right of=asm] {
  \nodepart{one} BEAM bytecode\\ ~
  \nodepart{two} \texttt{.beam}
};
\node (functional) [wrapper, fit={(source) (core)}, label=below:{\emph{Functional}}] {};
\node (imperative) [wrapper, fit={(asm) (bin)}, label=below:{\emph{Imperative}}] {};
\path [line] (source) -> (core);
\path [line] (core) -> (asm);
\path [line] (asm) -> (bin);
\end{tikzpicture}

\caption{Serializable representations along the path of a valid module
through the passes of the Erlang/OTP compiler. The symbol
\optimising{} marks the formats on which the compiler performs crucial
optimisations.}

\label{fig:serializable-representations}
\end{figure}

The following subsections detail how to use the Erlang/OTP
infrastructure to go from Erlang source code to BEAM bytecode, how to
run BEAM bytecode, and how to generate and pass in code in, and
further detail the aforementioned intermediate representations.

\subsection{Compiling Erlang source code to BEAM bytecode}

Erlang source code can be transformed into BEAM bytecode using the
Erlang compiler in Erlang/OTP. There are two entry-point binaries,
which we can use to interface with the compiler:

\begin{enumerate}

\item the Erlang compiler binary, \texttt{erlc}; and

\item the Erlang emulator and/or interactive shell, \texttt{erl}.

\end{enumerate}

For instance, assume that we have a file called \texttt{m.erl},
containing the source code of an Erlang module. We can compile it to
an \texttt{m.beam}, containing BEAM bytecode for that module, using
either of the ways shown in Figure \ref{fig:erlc}. These approaches
are non-interactive, in the sense that code is passed in, and passed
out via files, rather than standard input/output. Warnings and errors
however, are still written to standard output/error, in a
human-readable fashion.

\begin{figure}[ht!]
\begin{minipage}[t]{0.5\textwidth}
\begin{lstlisting}
$ erlc m.erl
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.5\textwidth}
\begin{lstlisting}
$ erl -compile m.erl
\end{lstlisting}
\end{minipage}%
\caption{A couple ``non-interactive'' ways to compile an Erlang
module.}
\label{fig:erlc}
\end{figure}

In either case, the work is done by the \texttt{compile} library in
Erlang/OTP. This means that an Erlang module can also be compiled in a
more interactive fashion, using \texttt{erl}. The potential benefit of
this approach is that warnings and errors can remain in a more
machine-readable format (i.e., Erlang terms, not bytes on a stream).
This approach is illustrated in Figure \ref{fig:erl-c}.

\begin{figure}[ht!]
\begin{minipage}[t]{0.2\textwidth}
\begin{lstlisting}
$ erl
...
1> R = c(m).
2>
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.35\textwidth}
\begin{lstlisting}
$ erl
...
1> R = compile:file(m).
2>
\end{lstlisting}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{lstlisting}
$ erl
...
1> R = compile:file(m, [return]).
2>
\end{lstlisting}
\end{minipage}%
\caption{A couple ``interactive'' ways to compile an Erlang module.
The variable \texttt{R} will contain information on the outcome of the
compilation. \texttt{c} is a simple variant of \texttt{compile:file}.
The right-most example will actually yield errors and warnings (if
any) as Erlang terms, inside the compound term \texttt{R}.}
\label{fig:erl-c}
\end{figure}

\subsection{Running BEAM bytecode}

Once compiled, the module code can be executed using \texttt{erl}.
Again, there is a non-interactive, and an interactive option.

\subsection{Core Erlang}

Core Erlang\cite{erlang:erlang:2001:core-intro, erlang:2004:core-spec}
is an intermediate language, designed specifically to enable the
Erlang/OTP compiler to leverage the compiler optimisations afforded by
a functional programming language, while not prohibiting the evolution
of either the Erlang syntax and semantics, or its execution
environment.

Although there exists a so-called ``Core Erlang 1.0.3 language
specification''\cite{erlang:2004:core-spec}, it hasn't been updated
since 2004. At the same time, the module in Erlang/OTP that implements
Core Erlang (i.e., \texttt{lib/compiler/src/cerl.erl}), has seen some
non-trivial modifications since 2004\footnote{At least, since 2007;
the source code available for download at \url{erlang.org} only goes
back to 2007. You will find a diff between the 2007 and 2019 versions
of \texttt{cerl} in \texttt{appendix/cerl.diff}.}. In particular,
associative maps were added as a built-in in Erlang in
2014\footnote{See
\url{http://erlang.org/download/otp_src_17.0.readme}.}; they made into
Core Erlang in 2013\footnote{See Erlang/OTP commit ID
\texttt{d4379900790b37498fefa237cd03ab2d5781bb6a}.}. Core Erlang
however, remains a high-level, strict, dynamically-typed, functional
language, with pattern-matching and concurrency primitives.  It bears
resemblance to Erlang, but lacks some of its high-level conveniences.

The optimisations performed by the Erlang/OTP compiler on Core Erlang
code are listed in Table \longref{tbl:core-erlang-optimisations}.
Overall, it performs function inlining, constant folding, and common
subexpression elimination.

\begin{table}[ht!]
\centering
\def\arraystretch{1.2}
\begin{tabular}{rlr}
\textbf{Order} & \textbf{Module} & \textbf{Line-count} \\\cline{2-3}
1 & \texttt{sys\_core\_inline} & 217 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Basic inlining of small functions.%
    }} \\\cline{2-3}
2 & \texttt{sys\_core\_fold} & 3276 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Constant folding.%
    }} \\\cline{2-3}
3 & \texttt{cerl\_inline} & 2834 \\
  & \multicolumn{2}{l}{\quad\emph{%
      More elaborate inlining, in lieau of Waddel \&
      Dybvig\cite{erlang:sas:1997:inlining}.%
    }} \\\cline{2-3}
4 & \texttt{sys\_core\_fold} & 3276 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Fold constants once more after aggressive inlining.%
    }} \\\cline{2-3}
5 & \texttt{sys\_core\_alias} & 307 \\
  & \multicolumn{2}{l}{\quad\emph{%
      A limited form of common subexpression elimination.%
    }} \\\cline{2-3}
6 & \texttt{sys\_core\_bsm} & 93 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Optimize patterns using Erlang's bit syntax.%
    }} \\\cline{2-3}
\end{tabular}
\caption{The optimisations performed on Core Erlang by the Erlang/OTP
compiler. All modules are located in \texttt{lib/compiler/src}, and
are called from the \texttt{compile} module, in the given order.
Line-counts are from Erlang/OTP version 22.1; they give  a rough idea
of where the majority of the work happens.}
\label{tbl:core-erlang-optimisations}
\end{table}

\texttt{erlc} can produce both optimised and unoptimised Core Erlang
code. For this, it accepts either the arugment \texttt{+to\_core}, or
\texttt{+to\_core0}, respectively. In either case, a \texttt{.core}
file is produced.

If you pass in a \texttt{.core} file to \texttt{erlc}, it will first
attempt to optimise the Core Erlang code, before moving on to the next
compiler stages. \texttt{erlc} takes the optional \texttt{+from\_core}
argument, to explicitly indicate that the incoming file is a Core
Erlang file. However, this argument is assumed for a \texttt{.core}
file.

\subsection{BEAM Assembly}

BEAM is a garbage collected, reduction counting, non-preemptive,
register-based virtual machine. These characteristics are portrayed in
the BEAM instruction set, and hence the BEAM assembly.

BEAM really has two instruction sets, a generic, and a specific one.
The purpose of this distinction is to enable innovation at the virtual
machine level, without introducing compiler-facing changes to the
instruction set. In particular, the generic instruction set is less
subject to change than the specific one. This is similar to the use of
microcode in Computer Architecture.

The file \texttt{lib/compiler/src/genop.tab} declares the generic
instruction set. The file uses a small, custom, human-readable,
domain-specific language. The Perl script,
\texttt{erts/emulator/utils/beam\_makeops}, uses the declarations to
generate both Erlang code for the compiler, and C code for the Erlang
runtime. This language, along with the Perl script, and the design
decisions behind them, are documented in
\texttt{erts/emulator/internal\_doc/beam\_makeops.md}\footnote{See
also \url{http://erlang.org/doc/apps/erts/beam_makeops.html} for the
most recent, typeset version of the file.}.

Most notably, the compiler is not made aware of the specific BEAM
instructions; the runtime takes care of transforming generic
instructions into specific ones. Hence, BEAM assembly is a
textualization of the generic BEAM instruction set, and all compiler
optimisations are done on generic BEAM instructions.

The optimisations performed by the Erlang/OTP compiler on BEAM
assembly, are listed in Table \longref{tbl:asm-optimisations}.
Overall, it performs function inlining, constant folding, and common
subexpression elimination.

\texttt{erlc} will produce optmized BEAM assembly, if you pass in the
argument \texttt{+to\_asm}, or equivalently, \texttt{-S}. If you pass
in a \texttt{.S} file, the argument \texttt{+from\_asm} is assumed,
and the compiler proceeds to generate BEAM bytecode.

\begin{table}[ht!]
\centering
\def\arraystretch{1.2}
\begin{tabular}{rlr}
\textbf{Order} & \textbf{Module} & \textbf{Line-count} \\\cline{2-3}
1 & \texttt{beam\_block} & 153 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Merge adjacent blocks; optimisations around blocks.%
    }} \\\cline{2-3}
2 & \texttt{beam\_except} & 255 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Rewrite some \texttt{erlang:error} calls to spec. instructions.%
    }} \\\cline{2-3}
3 & \texttt{beam\_jump} & 736 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Remove unreachable code, duplicate code; optimise jumps.%
    }} \\\cline{2-3}
4 & \texttt{beam\_peep} & 219 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Peep-hole optimisations; removing redundant tests.%
    }} \\\cline{2-3}
5 & \texttt{beam\_clean} & 167 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Cleanp up (e.g., remove unused labels and functions).%
    }} \\\cline{2-3}
6 & \texttt{beam\_trim} & 467 \\
  & \multicolumn{2}{l}{\quad\emph{%
      Stack trimming.
    }} \\\cline{2-3}
\end{tabular}
\caption{The optimisations performed on BEAM assembly by the
Erlang/OTP compiler. All modules are located in
\texttt{lib/compiler/src}, and are called from the \texttt{compile}
module, in the given order.  Line-counts are from Erlang/OTP version
22.1; they give  a rough idea of where the majority of the work
happens.}
\label{tbl:asm-optimisations}
\end{table}
